Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-04 paper/response.txt

We thank the reviewer for a timely report, and apologize for the delay
in our response. We have performed addition simulations (at lower
resolution) to address some of the reviewer's questions. The revisions
to the manuscript are made in bold, and further explained below.

- §§4 and 5: would the results of a simulation with a very short
  cooling time (beta << 1) be the same as those of a simulation with a
  locally isothermal equation of state and same unperturbed sound
  speed? 
  ********************************************************************
  We have performed addtional (lower resolution) simulations with
  beta=0.01 to compare with a locally isothermal disks, which we obtain 
  from our numerical models by setting gamma=1.01. We find a slightly
  smaller linear growth rate in the locally isothermal case. However,
  the vortex formation and evolution are similar. We    
  obtain longer gap-edge vortex lifetimes than beta=0.1 (at the same
  resolution) consistent with our double-peaked lifetime result for
  high resolution simulations.

@@

  ********************************************************************

- §4.4: why does the vortex’s perturbed density decrease on the long
  term? I suppose a related question is: what controls the long-term
  evolution of the vortex’s potential vorticity or entropy-modified
  potential vorticity? (whichever quantity is relevant). 
  ********************************************************************
  In theory a vortex should live forever in 2D inviscid flow. In
  practice, however, there is numerical viscosity which we suspect is
  responsible for the long term decay. As the vortex weakens
  (elongates) its area increases and its surface density drops. In
  addition, for beta=0.1, 0.5, 1.0, for which the initially-formed
  vortex is stronger, we observe the appearence of spiral waves
  associated with the vortex, which may contribute to the decay. It
  can be seen in Fig. 4 that these cases experience a larger decay
  rate.

@@

  ********************************************************************

- §5.2-5.3: here again I’d like to have more details on the vortex
        evolution and decay: 	
        . Do the authors see the vortex migrate? I did a simulation
        with the authors’ setup and a rather long cooling time
        (beta_tilde ~ 8) where I see the vortex moving from ~6.5 rh to
        ~5.3 rh relative to the planet’s orbital radius before the
        vortex starts decaying (rh = Hill radius). The time when the
        vortex starts decaying actually corresponds to a rather sudden
        increase in the planet-vortex distance, which goes up to ~6.8
        rh until the vortex dies off completely. If the authors get a
        similar migration behaviour in their simulations, maybe this
        could give further insight into the vortex evolution (vortex
        getting stronger during the quasi steady state before fading
        away). The connection between migration and the appearance of
        spiral waves launched by the vortex could also be made. 
	********************************************************************
        We thank the reviewer for raising the issue of vortex
        migration. We have re-examined our simulation results. We
        indeed observe vortex migration but our result is somewhat
        different to that of the reviewer.

 	We do not see significant vortex migration during quasi-steady
 	state. This may be due to the fact that the 	  
 	vortex forms at a surface density maximum, which is able to
 	trap vortices (Paardekooper et al 2010). Thus, the growth
 	during quasi steady state is not due to vortex migration.

@@

	We remark that for a vortex formed at a surface density
 	maximum (e.g. from the RWI), that subsequent perturbations may
 	lead to inwards or outwards vortex migration, depending on
 	what side of the maximum the vortex is displaced to
 	(Paardekooper et al 2010).

	Although we do not understand the discrepency between our
	results and the reviewer's, we do agree with the reviewer
	that vortex migration probably plays a role in determining
	the overall vortex lifetime (it may affect the duration of the
	quasi-steady state and/or the decay timescale). We now mention
	this in the discussions and caveats.

        ********************************************************************

        . During the quasi steady state the vortex grows as a result
        of continuous vortensity generation upon crossing the planet’s
        wake. What about possible turbulent diffusion of potential
        vorticity if the planet forms in a region of the disc with
        non-negligible ‘turbulent viscosity’? Have the authors done
        simulations with a non vanishingly small viscosity? 
        ********************************************************************
	Since a low-viscosity disc is needed for the RWI to operate,
        we have chosen to focus on invisicd simulations. This also
        helps to minimize complications from vortex decay directly due to a physical
        viscosity (which also contributes to heating).

### 2015-03-05 paper/results.tex
Final revision
Additonal Data Points

\subsection{Axisymmetric stability}
The initial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.45 \, \Omega^2(r_p)$
%$\sim 2\times10^{-4}$ (code units)
 for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency at the outer gap edge is
$N\sim 10^{-2} \, \Omega(r_p)$} 
% $N\sim 10^{-5}$,
decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

@@

We now examine vortex lifetimes as function of the imposed cooling
times. We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
% to reach $10\%$ of its max overdensity value
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
%{\bf correct interpretation?}
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 1300P_0$ for $\tilde{\beta}=0.1$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximises at $\tilde\beta=5.0$ with 
$t_{\mathrm{life}}\approx 1525P_0$.  The other timescales $t_\mathrm{diss}$ and
$t_\mathrm{Mach}$ are less complicated, but still display a
non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
$\tilde{\beta}=1$. 
% his is in comparison with the timescale of intial vortex dissipation which has 
% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
%non-monotonic dependence for all timescales

>>>>

\subsection{Axisymmetric stability}
The initial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.45 \, \Omega^2(r_p)$
%$\sim 2\times10^{-4}$ (code units)
 for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency at the outer gap edge is
$N\sim 0.1 \, \Omega(r_p)$} 
% $N\sim 10^{-5}$,
decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

@@

We now examine vortex lifetimes as function of the imposed cooling
times. {\bf For completeness additional simultions were done for $\tilde\beta=0.01,0.25,0.75,2.5,7.5$. }
 We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
% to reach $10\%$ of its max overdensity value
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
%{\bf correct interpretation?}
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximises at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
$t_\mathrm{Mach}$ are less complicated, but still display a
non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
$\tilde{\beta}=1$. 
% his is in comparison with the timescale of intial vortex dissipation which has 
% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
%non-monotonic dependence for all timescales

### 2015-03-05 paper/summary.tex
Additonal Data Points

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=5.0$, which has
the longest vortex lifetime, that  
$h\approx0.06$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.

>>>>

We remark that a non-monotonic dependence of the vortex lifetime was
also reported by \cite{fu14}, who performed isothermal disc-planet
simulations with different values of the   
disc aspect ratio. In their simulations the optimum aspect ratio is
$h=0.06$. In our simulations, $h$ is a dynamical
variable, but by analysing the region where the vortex is located
($r-r_p\in[2,10]r_h$), % {\bf define this region}
we find for a dimensionless cooing timescale of $\tilde\beta=2.5$, which has
the longest vortex lifetime, that  
$h\approx0.058$ on average.  
%{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore consistent with \citep{fu14}.
